\( \tool \): JavaScript IR-based Semantics Extraction Toolchain

### 2020-04-12 ase20/compiler.tex
[ASE] Updated Section 4.

\section{Algorithm Compiler}\label{sec:compiler}
\begin{figure}
  \centering
  \includegraphics[width=0.48\textwidth]{img/algo_compiler.png}
  \caption{Overall structure of {\sf Algorithm Compiler}}
  \label{fig:algo-compiler}
\end{figure}

@@

\subsection{Tokenizer}
Before compiling abstract algorithms, {\sf Tokenizer} first tokenizes
each abstract algorithm into a list of tagged tokens.  An algorithm
consists of ordered steps, and a step may contain sub-steps as well.
For example, the \textbf{\small ToPrimitive} algorithm in
Figure~\ref{fig:to-primitive}(a) has three steps and the second step
has seven sub-steps.  Moreover, the tokens of each step have their own
HTML tags and each tag has a meaning.  For instance, \code{<var>}
denotes variables, \code{<emu-grammar>} denotes productions, and
\code{<code>} denotes ECMAScript code.
% \[
%   \begin{array}{c|l}
%     \text{HTML tags} & \text{meanings}\\\hline\hline
%     \code{<var>} & \text{variables}\\\hline
%     \code{<emu-grammar>} & \text{productions}\\\hline
%     \code{<emu-nt>} & \text{non-terminal syntax}\\\hline
%     \code{<code>} & \text{ECMAScript codes}\\\hline
%     \code{<emu-const>} & \text{constant values}\\\hline
%     \code{<emu-val>} & \text{values}\\\hline
%     \code{<ol>} & \text{ordered sub-steps}\\\hline
%     \code{<ul>} & \text{unordered sub-steps}\\\hline
%     \code{<sup>} & \text{superscripts}\\\hline
%     \text{otherwise} & \text{simple texts}\\\hline
%   \end{array}
% \]
We keep such HTML tag information for each token to generate more precise
{\sf Compile Rules}.  For example, because if and only if a token has
a tag \( \code{<var>} \), it is a parameter or a local variable, we can
construct {\sf Compile Rules} that precisely distinguish identifiers
from other components.

The {\sf Tokenizer} module recognizes the structure of each step,
and divides the step into a sequence of tagged tokens.
If an HTML element has an explicit tag, it is converted to a single token
with the tag.  Otherwise, it is divided into multiple tokens and each token
becomes a sequence of alphanumeric characters or a single non-alphanumeric
character.  For example, in the \textbf{\small ToPrimitive} algorithm,
\( \textbf{\code{"default"}} \) is a single token with the HTML tag \( \code{<code>} \)
and \( \code{@@toPrimitive} \) is divided into three text tokens
\( \code{@} \), \( \code{@} \), and \( \code{toPrimitive} \).

Moreover, {\sf Tokenizer} flattens a structured step to a single token
list to handle multi-step statements easily.  Some statements in
abstract algorithms consist of multiple steps.  For example, the
\( \code{if}\!-\!\code{then}\!-\!\code{else} \) statement often consists of
two steps: one for the \( \code{then} \)-branch and another
for the \( \code{else} \)-branch.  To treat them as a linear structure,
we introduce three special tokens to break down structured algorithms:
\( \tend \) denotes the end of a single step, and \( \tin \) and
\( \tout \) denote the start and the end of nested steps, respectively.
For example, the \textbf{\small ToPrimitive} algorithm is tokenized as follows.
\[
  \begin{array}{l}
    \code{Assert} \cdots \tend\\
    \code{If} \cdots \tin \code{If} \cdots \tend
    \cdots \code{Return} \cdots \tend \tout \tend\\
    \code{Return} \cdots \tend\\
  \end{array}
\]

@@

\subsection{Token List Parser}
Parsing rules determine the token list parser, and each parsing rule
is written in extended Scala parser combinators.  We modified the
meaning of the alternative composition operator ( \( ||| \) ) to collect
all the longest matched results.  If a parser detects a step that
cannot be parsed with given parsing rules, it reports the step into
{\sf Fault Localizer}.  Even when a parser parses a step successfully,
it also reports the step if it can parse the step in multiple ways.
We use Packrat parsing~\cite{packrat} for parsing rules, which
support left recursion.

We provide two kinds of basic token parsers: tag-based parsers and
content-based parsers.  A tag-based parser just checks whether the next
token has a given tag.  For example, the tag-based parser \( \code{varT} \)
checks whether the next token has the tag \( \code{<var>} \).  A
content-based parser checks whether the next token is a text token and
its content passes a given condition.  For example, the string literal
\( \code{"Let"} \) denotes a content-based parser that checks whether
the next token is a text token with the content \( \code{Let} \).
We also provide two content-based parsers \( \code{word} \) and
\( \code{number} \) that check whether the content of the next token
consists of only alphabets or numbers, respectively.  In parsing
rules, all the helper functions defined in Scala parser combinators
are available.  For instance, the helper function \( \code{repsep(p, q)} \)
generates a new parsing rule that denotes zero or more repetition of the
parsing rule \( \code{p} \) using another parsing rule \( \code{q} \) as
a separator.

Consider the following example parsing rule, which is a simplified
version for the step 2-e-i of the \textbf{ToPrimitive} algorithm.
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = "Let" ~ varT ~ "be" ~ Expr ~ "." ^^ ...
// expressions
val Expr =
  // identifiers        // return if abrupt
  varT      ^^ ... |||  "?" ~ Expr      ^^ ... |||
  // function calls
  word ~ "(" ~ repsep(Expr, ",") ~ ")"  ^^ ... |||
  // lists
  "<<" ~ repsep(varT, ",") ~ ">>"       ^^ ...
\end{lstlisting}
The \( \code{Stmt} \) compile rule describes how to compile statements
with one parsing rule and its conversion rule is omitted for now for
brevity.  The \( \code{Expr} \) compile rule describes how to compile
expressions with four parsing rules.  A token parser with the above
rules parses the step 2-e-i of \textbf{ToPrimitive} to the following
token AST:
\begin{center}
\includegraphics[width=0.5\textwidth]{img/token_ast.png}
\end{center}

\subsection{Token AST Converter}
Conversion rules describe how to generate an \( \ires \) function for a
given token AST.  Each conversion rule is defined with its
corresponding parsing rule.  For basic token parsers, their conversion
rules always return the string values of the contents in parsed tokens.
For example, the following conversion rules are the omitted parts
in the previous example for the step 2-e-i of \textbf{ToPrimitive}:
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = ... ^^ { case _ ~ x ~ _ ~ y => ILet(x, y) }
// expressions
val Expr =
  // identifiers   // return if abrupt
  ... ^^ EId |||   ... ^^ { _ ~ e => EAbrupt(e) } |||
  // function calls
  ... ^^ { case x ~ _ ~ y ~ _ => ECall(x, y) }    |||
  // lists
  ... ^^ { case _ ~ x ~ _ => EList(e) }
)
\end{lstlisting}
The conversion rule of the compile rule \( \code{Stmt} \) uses only
the second and fourth sub trees and converts them to \( \ires \)
instructions.  The second sub-tree is converted to a string value
because it was constructed via the basic token parser \( \code{varT} \).
The fourth sub-tree is converted to an \( \ires \) expression
because the conversion rules of the \( \code{Expr} \) compile rule
construct \( \ires \) expressions.  Then, the conversion rule of
\( \code{Stmt} \) constructs the lexical binding instruction \( \code{ILet} \)
with the string value and the \( \ires \) expression.  In this way,
the step 2-e-i of \textbf{ToPrimitive} is converted to the following
\( \ires \) instruction:
\begin{lstlisting}[style=ires]
let result = ? (Call exoticToPrim input (new [hint]))
\end{lstlisting}

We define \( \ires \) to represent abstract algorithms as its
functions with the following design choices:

\begin{itemize}[leftmargin=0.5cm]
\item \textbf{Dynamic typing:} Because each variable in abstract
algorithms is not statically typed, variables do not have their own
static types while each value of \( \ires \) has its dynamic type.

@@

\item \textbf{Primitive values:} \( \ires \) supports ECMAScript
primitive values except ``symbols'' because symbols can be represented
as singleton objects.  Also, \( \ires \) provides the unique \(
\code{absent} \) value to represent the absence of parameters.  For
example, when the optional second parameter \( \code{PreferredType} \)
of \textbf{ToPrimitive} in Figure~\ref{fig:to-primitive}(a) is absent,
the parameter has the \( \code{absent} \) value.

\item \textbf{Abstract data types:} \( \ires \) supports only three
abstract data types: \( \code{Record} \) for mappings from some values
to other values, \( \code{List} \) for sequential data, and
\( \code{Singleton} \) for singleton data.  For example, ECMAScript
environment records are represented as \( \code{Record} \)
from string values into addresses that represent other types of
\( \code{Record} \) for bindings.  We also represent constants
in the ECMAScript specification such as \( \code{empty} \)
and \( \code{normal} \) as \( \code{Singleton} \) data types
in \( \ires \).
\end{itemize}

\subsection{Fault Localizer}\label{sec:localizer}
\begin{algorithm}[t]
  \caption{Fault Localization}\label{alg:fault}
  \begin{algorithmic}[1]
    \Function{FaultLocalize}{$T,n$}\Comment{$T$: failed step, $n$: size}
    \State $S = \varnothing$
    \For{$i = 0$ to $n - 1$}
    \For{$j = i$ to $n - 1$}
    \State $T'$ = a copy of $T$
    \State Substitute $T'[i..j]$ with $\tstar$
    \If{Parse($T'$) contains $\code{Dummy(}x \code{)}$ for some $x$}
    \State Insert $(x, i, j)$ into $S$
    \EndIf
    \EndFor
    \EndFor
    \While{$s = (x, i, j) \in S$}
    \If{$\exists (y, p, q) \in S$ s.t. $i \leq p \wedge q \leq j$}
    \State Remove $s$ from $S$
    \EndIf
    \EndWhile
    \State \Return $S$
    \EndFunction
  \end{algorithmic}
\end{algorithm}

To diagnose root causes of parsing failures, we support a \textsf{Fault Localizer},
which takes not-yet-convertible steps dubbed ``failed steps'' as arguments.
The basic idea is to check whether we can parse the failed algorithm step
with the assumption that a specific sub-sequence of the
failed step is successfully parsed.  We define a \textit{rule suggestion}
as such a sub-sequence of failed steps with the name of an existing compile rule.
\textsf{Fault Localizer} refines candidates of rule suggestions into the
most specific ones. It helped us when we update existing
compile rules for new versions of the specification.

The fault localization algorithm is shown in Algorithm~\ref{alg:fault}.
For a given failed step $T$ and the size of a failed input $n$, it
attempts to find a set of the most specific rule suggestions.  It
substitutes a specific sub-sequence \( T[i..j] \) of \( T \) with the
special token \( \tstar \), which represents some sub-sequence of a
failed step.  It collects a set of possible rule suggestions, and then
removes more general ones than existing suggestions.  To support this
process, we added a parsing rule  \( \code{starT} \) to each compile
rule as follows:
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = ... ||| starT ^^ { case x => Dummy("Stmt") }
// expressions
val Expr = ... ||| starT ^^ { case x => Dummy("Expr") }
\end{lstlisting}
where \( \code{starT} \) is a special basic token parser that just
checks whether a given token is the special token \( \tstar \). 
Its corresponding conversion rule is a function that returns a dummy
\( \ires \) component \( \code{Dummy} \) with a string value that
represents the name of its compile rule.

For example, let us assume that the \( \code{Expr} \) compile rule
does not have the alternative for lists:
\begin{lstlisting}[style=myScalastyle]
  "<<" ~ repsep(varT, ",") ~ ">>" ^^
       { case _ ~ x ~ _ => EList(e) } // lists
\end{lstlisting}
Then, the step 2-e-i of \textbf{ToPrimitive} becomes a failed step.
When we use the fault localization algorithm with the failed step and
its size 15, it first finds one rule suggestion for statements and four
rule suggestions for expressions as follows:
\[
\footnotesize
\! \! \! \!
  \begin{array}{c}
    (\scode{"Stmt"}, 0, 14), \;
    (\scode{"Expr"}, 3, 13), \;
    (\scode{"Expr"}, 4, 13), \;
    (\scode{"Expr"}, 6, 12), \;
    (\scode{"Expr"}, 10, 12)
  \end{array}
\]
In the \textbf{\small while} loop on lines 9--11, because the last one
is the most specific sub-sequence, it removes the other four
rule suggestions. Finally, \textsf{Fault Localizer} suggests to define
another expression compile rule for the token list \hint{}.

>>>>

\section{Algorithm Compiler}\label{sec:compiler}
\begin{figure}
  \centering
  \includegraphics[width=0.48\textwidth]{img/algo-compiler.pdf}
  \caption{overall structure of {\sf algorithm compiler}}
  \label{fig:algo-compiler}
\end{figure}

@@

\subsection{Tokenizer}
Before compiling abstract algorithms, {\sf Tokenizer} first tokenizes each
abstract algorithm into a list of tagged tokens.  An algorithm consists of
ordered steps, and a step may contain sub-steps as well.  For example, the
\textbf{\small Evaluation} abstract algorithm in
Figure~\ref{fig:array-literal-eval}(a) has seven steps.  Moreover, the tokens of
each step have their own HTML tags and each tag has a meaning.  We keep such
HTML tag information for each token to construct more precise {\sf Compile
Rules}. If an HTML element is just a text without any explicit tags, it is
divided into multiple tokens and each token becomes a sequence of alphanumeric
characters or a single non-alphanumeric character.  For example, in the
\textbf{\small Evaluation} algorithm, \( \textbf{\code{"length"}} \) is a single
token with the HTML tag \( \code{<code>} \) and \( \code{Perform Set(} \) is
divided into three text tokens \( \code{Perform} \), \( \code{Set} \), and \(
\code{(} \).

Moreover, {\sf Tokenizer} flattens a structured step to a single token
list to handle multi-step statements easily.  Some statements in
abstract algorithms consist of multiple steps.  For example, the
\( \code{if}\!-\!\code{then}\!-\!\code{else} \) statement often consists of
two steps: one for the \( \code{then} \)-branch and another
for the \( \code{else} \)-branch.  To treat them as a linear structure,
we introduce three special tokens to break down structured algorithms:
\( \tend \) denotes the end of a single step, and \( \tin \) and
\( \tout \) denote the start and the end of nested steps, respectively.
For example, the following left abstract algorithm is tokenized to the right
token list.
\[
  \begin{array}{lcl}
    \code{1. A}\\
    \code{2. B} &\Longrightarrow& \code{A} \tend \code{B} \tin \code{C} \tend \tout \tend\\
    \code{\ \ a. C}\\
  \end{array}
\]

@@

\subsection{Token List Parser}
The token list parser is defined with \textit{parsing rules} written in extended
Scala parser combinators. We modified the meaning of the alternative composition
operator ( \( | \) ) to collect all the longest matched results. If a parser
detects a step that cannot be parsed or parsed in multiple ways, it reports the
step with the parsing results.  We use Packrat parsing~\cite{packrat} for
parsing rules, which support left recursion.

We provide two kinds of basic parsing rules: \textit{tag-based rules} and
\textit{content-based rules}.  A tag-based rule just checks whether the next
token has a given tag.  For example, the tag-based parser \( \code{varT} \) and
\( \code{codeT} \) check whether the next token has the tag \( \code{<var>} \)
and \( \code{<code>} \), respectively.  A content-based parser checks whether
the next token is a text token and its content passes a given condition.  For
example, the string literal \( \code{"Perform"} \) denotes a content-based
parser that checks whether the next token is a text token with the content \(
\code{Perform} \).  We also provide two content-based parsers \( \code{word} \)
and \( \code{number} \) that check whether the content of the next token
consists of only alphabets or numbers, respectively.  In parsing rules, all the
helper functions defined in Scala parser combinators are available.  For
instance, the helper function \( \code{repsep(p, q)} \) generates a new parsing
rule that denotes zero or more repetition of the parsing rule \( \code{p} \)
using another parsing rule \( \code{q} \) as a separator.

Consider the following example parsing rule for the step 5 of the
\textbf{Evaluation} algorithm in Figure~\ref{fig:array-literal-eval}(a).
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = "Perform" ~ Expr ~ "." ^^ ...
// expressions
val Expr =
  // codes        // false literal
  codeT    ^^ ... |  "false"           ^^ ... |
  // variables       // additions
  varT     ^^ ... |  Expr ~ "+" ~ Expr ^^ ... |
  // function calls
  word ~ "(" ~ repsep(Expr, ",") ~ ")" ^^ ...
\end{lstlisting}
We omit the conversion rule for each compile rule for the brevity.  The \(
\code{Stmt} \) compile rule  describes how to compile statements with a single
parsing rule, and the \( \code{Expr} \) compile rule describes how to compile
expressions with five parsing rules.  A token parser with the above rules parses
the step 5 of \textbf{Evaluation} to the following token AST:
\begin{center}
  \includegraphics[width=0.5\textwidth]{img/token-ast.pdf}
\end{center}

\subsection{Token AST Converter}
\textit{Conversion rules} describe how to generate an \( \ires \) function for a
given token AST.  Each conversion rule is defined with its corresponding parsing
rule.  For basic parsing rules, their conversion rules always return the string
values of the contents in parsed tokens.  For example, the following conversion
rules are the omitted parts in the previous example for the step 5 of
\textbf{Evaluation}:
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = ... ^^ { case _ ~ e ~ _ => IExpr(e) }
// expressions
val Expr =
  // codes      // false literal
  ... ^^ EStr | ... ^^ { _ => EBool(false) } |
  // variables  // additions
  ... ^^ EId  | ... ^^ { case x ~ _ ~ y => EAdd(x, y) } |
  // function calls
  ... ^^ { case x ~ _ ~ y ~ _ => ECall(x, y) }
)
\end{lstlisting}
The conversion rule of the \code{Stmt} compile rule uses only the second sub
tree and constructs an \code{IExpr} \( \ires \) instruction.  For the second
sub-tree, the conversion rule of the fourth \code{Expr} compile rule is
applied. It constructs \code{ECall} \( \ires \) expression with the string value
of the first sub-tree and the sequence of the expressions of the third sub-tree.
In this way, the step 5 of \textbf{Evaluation} is converted to the following \(
\ires \) instruction:
\begin{lstlisting}[style=ires]
(Set array "length" (ToUint32 (+ padding len)) false)
\end{lstlisting}

We define \( \ires \) to represent abstract algorithms as its
functions with the following design choices:
\begin{itemize}[leftmargin=0.5cm]
\item \textbf{Dynamic typing:} Because each variable in abstract
algorithms is not statically typed, variables do not have their own
static types while each value of \( \ires \) has its dynamic type.

@@

\item \textbf{Primitive values:} \( \ires \) supports ECMAScript primitive
values except ``symbols'' because symbols can be represented as singleton
objects.  Also, \( \ires \) provides the unique \( \code{absent} \) value to
represent the absence of parameters.  For example, when the optional second
parameter \textit{Elision} of \textbf{Evaluation} in
Figure~\ref{fig:array-literal-eval}(a) is absent, the parameter has the \(
\code{absent} \) value.

\item \textbf{Abstract data types:} \( \ires \) supports only three abstract
data types: \( \code{Record} \) for mappings from some values to other values,
\( \code{List} \) for sequential data, and \( \code{Symbol} \) for singleton
data.  For example, ECMAScript environment records are represented as \(
\code{Record} \) from string values into addresses that represent other types
of \( \code{Record} \) for bindings.
\end{itemize}
We define syntax of \( \ires \) consisting 15 kinds of instructions and 26 kinds
of expressions with the notation \( \irinst \) and \( \irexpr \), respectively.
Moreover, we also formally define their operational semantics with the form \(
\irstate \vdash \irinst \Rightarrow \irstate \) for instructions, and the form
\( \irstate \vdash \irexpr \Rightarrow (\irvalue, \irstate) \) for expressions
where \( \irvalue \) denotes a value and \( \irstate \) denotes a state.
However, we omit this formalization in this paper for brevity, instead, we
include it in the companion material\footnote{\inred{???}}.
